Surface correlation functions calculations: improving methodology for discrete images with novel exact continuous approach

### 2023-03-29 paper.tex
Fix typos and errors

We give an image of porous medium in the form of two- or three-dimensional bit
array $A$ as the input to our algorithm. In the first stage, we convolve the
input with high-pass filter to extract edges from the image. We use the
following filter kernel $H$:
\begin{equation}
  \begin{aligned}
    H_{ij} &= \frac{1}{6} \left[
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & -8 & 1 \\
        1 & 1 & 1
      \end{array}
      \right] \\
    H_{ijk} &= \frac{1}{18} \left[
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & 1 & 1 \\
        1 & 1 & 1
      \end{array} ;
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & -24 & 1 \\
        1 & 1 & 1
      \end{array} ;
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & 1 & 1 \\
        1 & 1 & 1
      \end{array}
      \right]
  \end{aligned}
  \label{eq:filter-3x3}
\end{equation}
Here symbol $;$ denotes concatenation along the third dimension. Edge extraction
is performed as follows:
\begin{align*}
  A'_{lm}  &= \left| \sum_i\sum_j A_{l-i, m-j}H_{ij} \right| \quad \text{for 2D} \\
  A'_{lmn} &= \left| \sum_i\sum_j\sum_k A_{l-i, m-j, n-k}H_{ijk} \right| \quad
  \text{for 3D}
\end{align*}
Therefore $A'$ has non-zero elements only in a small area around interface
between phases in $A$. This operation is convolution of a filter $H$ with an
image $A$ and can be performed using Fourier transform:
\begin{equation*}
  A' = H*A = F^{-1}[F[H] F[A]]
\end{equation*}
Here $F$ is a forward Fourier transform and $F^{-1}$ is its inverse.

@@

All in all we get the following algorithm for calculation of $F_{ss}$ function:
\begin{algorithmic}[1]
  \Procedure{surfsurf}{$A$}
  \State $A' \gets H*A$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $\hat{A'} \gets F[A']$
  \Comment{Apply DFT to $A'$.}
  \State \textbf{return} $F^{-1}[ |\hat{A}|^2 / N]$
  \Comment{$N$ is a number of elements in $A$}
  \EndProcedure
\end{algorithmic}

Algorithm for surface-void function $F_{sv}$ is similar:
\begin{algorithmic}[1]
  \Procedure{surfvoid}{$A$}
  \State $A' \gets A*H$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $V \gets \chi_V(A)$
  \Comment{Apply an indicator function for void phase element-wise to $A$.}
  \State $\hat{A'} \gets F[A']$
  \Comment{Apply DFT to $A'$.}
  \State $\hat{V} \gets F[V]$
  \Comment{Apply DFT to $V$.}
  \State \textbf{return} $F^{-1}[ \hat{A}\overline{\hat{V}} / N]$
  \Comment{$N$ is a number of elements in $A$}
  \EndProcedure
\end{algorithmic}

@@

For surface-surface-surface function $F_{sss}$ the former approach can also be
used, but it would be highly inefficient memory-wise because for three
dimensional case $F_{sss}$ is a function of six scalar coordinates which leads
to a very large output array. Instead of this approach we use a slower approach
which computes $F_{sss}$ at a specified point calculating a convolution using
its definition:
\begin{algorithmic}[1]
  \Procedure{surfsurfsurf}{$A$, $\bm{r_1}$, $\bm{r_2}$}
  \State $A' \gets A*H$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $A'_1 \gets shift(A', \bm{r_1})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{1i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $A'_2 \gets shift(A', \bm{r_2})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{2i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $M \gets A' \cdot A'_1 \cdot A'_2$
  \Comment{Multiplicate three arrays element-wise}
  \State \textbf{return} $\sum_{i} M_{i} / N$
  \Comment{Sum all elements in $M$ and divide by a number of elements in $A$}
  \EndProcedure
\end{algorithmic}
This algorithm can be vectorized and therefore benefits from the usage of
GPUs. Shift function in this algorithm must be a circular shift which
``rotates'' content of the array, if periodic boundary conditions are
desired. For our purpose we use $F_{sss}$ function in the form
$F_{sss}(r_1, r_2, \theta)$ and fix $\theta = \pi/2$. This leads us to a
sampling pattern shown on \cref{fig:Fsss-pattern}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/pattern.png}
  \caption[]{Sampling pattern used for calculation of $F_{sss}$ in our
    implementation is a right triangle with varying catets $r_1$ and
    $r_2$. Patterns can lie in planes $XY$ (as on the picture), $XZ$ or
    $YZ$. Black dots depict elements of $A'$}
  \label{fig:Fsss-pattern}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/experiment-setup.png}
  \caption[]{Left, center: images $A$ and $B$ used in the experiment for
    approximation of $1 / \sin \phi$. Right: Interfaces between phases are
    line segments which intersect each other at angle $\phi$. Shaded area is
    where $|(H*A)\cdot(H*B)|$ has non-zero values. Sum of all values in this
    area equals to $1 / \sin \phi$.}
  \label{fig:experiment-setup}
\end{figure}
Take two 2D images like the ones on \cref{fig:experiment-setup}. Boundaries
between black and white areas cross each other at the angle $\phi$. We apply a
edge detection filter to each image, multiply results element-wise and sum all
elements of the product. We want an obtained value to approximate
$1/\sin \phi$. As can be seen on \cref{fig:filter-comparison} $H'$ does this job
better than $H$. Why it's so important will be explained in the next section.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/filter-comparison.png}
  \caption[]{Estimation of $1/\sin\phi$ by filters $H$ and $H'$ in the
    experiment on \cref{fig:experiment-setup}.}
  \label{fig:filter-comparison}
\end{figure}

>>>>

We give an image of porous medium in the form of two- or three-dimensional bit
array $A$ as the input to our algorithm. In the first stage, we convolve the
input with high-pass filter to extract edges from the image. We use the
following filter kernel $H$:
\begin{equation}
  \begin{aligned}
    H_{ij} &= \frac{1}{6} \left[
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & -8 & 1 \\
        1 & 1 & 1
      \end{array}
      \right] \\
    H_{ijk} &= \frac{1}{18} \left[
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & 1 & 1 \\
        1 & 1 & 1
      \end{array} ;
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & -24 & 1 \\
        1 & 1 & 1
      \end{array} ;
      \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & 1 & 1 \\
        1 & 1 & 1
      \end{array}
      \right]
  \end{aligned}
  \label{eq:filter-3x3}
\end{equation}
Here symbol $;$ denotes concatenation along the third dimension. Edge extraction
is performed as follows:
\begin{align*}
  A'_{lm}  &= \left| \sum_i\sum_j A_{l-i, m-j}H_{ij} \right| \quad \text{for 2D} \\
  A'_{lmn} &= \left| \sum_i\sum_j\sum_k A_{l-i, m-j, n-k}H_{ijk} \right| \quad
  \text{for 3D}
\end{align*}
Therefore $A'$ has non-zero elements only in a small area around interface
between phases in $A$. This operation is convolution of a filter $H$ with an
image $A$ and can be performed using Fourier transform:
\begin{equation*}
  A' = |H*A| = |F^{-1}[F[H] F[A]]|
\end{equation*}
Here $F$ is a forward Fourier transform and $F^{-1}$ is its inverse.

@@

All in all we get the following algorithm for calculation of $F_{ss}$ function:
\begin{algorithmic}[1]
  \Procedure{surfsurf}{$A$}
  \State $A' \gets |H*A|$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $\hat{A'} \gets F[A']$
  \Comment{Apply DFT to $A'$.}
  \State \textbf{return} $F^{-1}[ |\hat{A}|^2 / N]$
  \Comment{$N$ is a number of elements in $A$}
  \EndProcedure
\end{algorithmic}

Algorithm for surface-void function $F_{sv}$ is similar:
\begin{algorithmic}[1]
  \Procedure{surfvoid}{$A$}
  \State $A' \gets |H*A|$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $V \gets \chi_V(A)$
  \Comment{Apply an indicator function for void phase element-wise to $A$.}
  \State $\hat{A'} \gets F[A']$
  \Comment{Apply DFT to $A'$.}
  \State $\hat{V} \gets F[V]$
  \Comment{Apply DFT to $V$.}
  \State \textbf{return} $F^{-1}[ \hat{A}\overline{\hat{V}} / N]$
  \Comment{$N$ is a number of elements in $A$}
  \EndProcedure
\end{algorithmic}

@@

For surface-surface-surface function $F_{sss}$ the former approach can also be
used, but it would be highly inefficient memory-wise because for three
dimensional case $F_{sss}$ is a function of six scalar coordinates which leads
to a very large output array. Instead of this approach we use a slower approach
which computes $F_{sss}$ at a specified point calculating a convolution using
its definition:
\begin{algorithmic}[1]
  \Procedure{surfsurfsurf}{$A$, $\bm{r_1}$, $\bm{r_2}$}
  \State $A' \gets |H*A|$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $A'_1 \gets shift(A', \bm{r_1})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{1i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $A'_2 \gets shift(A', \bm{r_2})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{2i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $M \gets A' \cdot A'_1 \cdot A'_2$
  \Comment{Multiplicate three arrays element-wise}
  \State \textbf{return} $\sum_{i} M_{i} / N$
  \Comment{Sum all elements in $M$ and divide by a number of elements in $A$}
  \EndProcedure
\end{algorithmic}
This algorithm can be vectorized and therefore benefits from the usage of
GPUs. Shift function in this algorithm must be a circular shift which
``rotates'' content of the array, if periodic boundary conditions are
desired. For our purpose we use $F_{sss}$ function in the form
$F_{sss}(r_1, r_2, \theta)$ and fix $\theta = \pi/2$. This leads us to a
sampling pattern shown on \cref{fig:Fsss-pattern}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/pattern.png}
  \caption[]{Sampling pattern used for calculation of $F_{sss}$ in our
    implementation is a right triangle with varying catets $r_1$ and
    $r_2$. Patterns can lie in planes $XY$ (as on the picture), $XZ$ or
    $YZ$. Black dots depict elements of $A'$}
  \label{fig:Fsss-pattern}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/experiment-setup.png}
  \caption[]{Left, center: images $A$ and $B$ used in the experiment for
    approximation of $1 / \sin \phi$. Right: Interfaces between phases are
    line segments which intersect each other at angle $\phi$. Shaded area is
    where $|(H*A)\cdot(H*B)|$ has non-zero values. Sum of all values in this
    area approximates $1 / \sin \phi$.}
  \label{fig:experiment-setup}
\end{figure}
Take two 2D images like the ones on \cref{fig:experiment-setup}. Boundaries
between black and white areas cross each other at the angle $\phi$. We apply an
edge detection filter to each image, multiply results element-wise and sum all
elements of the product. We want an obtained value to approximate
$1/\sin \phi$. As can be seen on \cref{fig:filter-comparison} $H'$ does this job
better than $H$. Why it's so important will be explained in the next section.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/filter-comparison.png}
  \caption[]{Estimation of $1/\sin\phi$ by filters $H$ and $H'$ in the
    experiment on \cref{fig:experiment-setup}.}
  \label{fig:filter-comparison}
\end{figure}
